Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Isothermal Stellar Envelopes

### 2020-01-22 writeups/Su_IGW_break.tex
Latest comments from Dong

\title[Internal Gravity Wave Breaking]{Physics of Tidal Dissipation:
Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar
Envelopes and Massive Stellar Binaries}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^2$
Dong Lai,$^1$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
}

@@

\pubyear{2019}

@@

\begin{abstract}
    In sufficiently compact white dwarf binaries, dynamical tides raise a train
    of internal gravity waves that propagate towards the surface and dissipate
    via nonlinear wave breaking. We perform 2D hydrodynamical simulations of
    this wave breaking in an incompressible, isothermal atmosphere. After an
    initial transient phase, we find that these waves generate a sharp
    transition, a critical layer, between the non-rotating core and
    synchronously rotating envelope. We find evidence that indefinite steepening
    of this critical layer is suppressed by the Kelvin-Helmholtz instability. We
    study the absorption and reflection of incident waves off the critical layer
    and provide analytical formulae describing its evolution. We infer
    necessary conditions to resolving momentum transfer within the critical
    layer. Finally, we speculate on the application of our model to tidal
    synchronization and heating in astrophysical systems.
\end{abstract}

@@

@@

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA) \citep{lisa}. They are also thought to
produce interesting optical transients such as underluminous supernovae
\citep{underlum}, Ca-rich fast transients \citep{carich}, and tidal novae
\citep{tidal_novae}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g. \citet{Ia0,webbink} or more recently
\citet{Ia1,Ia2}). While presently only a few tens of compact WD binaries are
known \citep{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundreds \citep{lsst_wd} \citep[results based on
\emph{Gaia}'s second data release have already begun to
appear][]{gaiaDD,gaiaDD2}, and the Large Synoptic Survey Telescope (LSST, first
light scheduled for 2020) will likely detect a few thousand more
\citep{lsst_wd}. These observations will significantly advance the understanding
of WD binaries and their evolution.

In spite of the broad importance of WD binaries, the evolution of these systems
prior to their final mergers is not well understood. Much of this uncertainty
comes from our imprecise understanding of tidal interactions, which play an
important role during a compact WD binary's inspiral \citep{fullerII}. Previous
studies have shown that these interactions manifest as tidal excitation of
internal gravity waves (IGW), waves in the WD fluid restored by the buoyancy
force due to density stratification \citep{fullerI}. As these waves propagate
outwards towards the WD surface, they grow in amplitude until they break, as do
ocean waves on a shore, and transfer both energy and angular momentum from the
binary orbit to the outer envelope of the WD \citep{fullerI,fullerII}.

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget \citep{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and simply brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares \citep{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking behaves differently than predictions
based in linear and weakly nonlinear theory \citep{winters1994,barker_ogilvie}.
Such fully nonlinear numerical simulations have not been performed for WDs.

In this paper, we perform numerical simulations of tidally-excited IGWs
undergoing wave breaking in an idealized WD fluid. The primary aim of the paper
is to thoroughly characterize the angular momentum transfer within a WD by
purely hydrodynamical effects. As a first approximation, we analyze the problem
in a 2D Cartesian geometry, and consider IGW propagating into a fluid initially
at rest. We find that, after an initial transient phase, the fluid develops two
distinct zones: (i) a lower zone that has no horizontal mean flow, and (ii) an
upper zone with significant horizontal mean flow.

These two zones are separated by a \emph{critical layer}. The interaction of IGW
and critical layers plays an important role in the tidal synchronization process
of stellar binaries \citep{zahn75,gn89} as well as in terrestrial phenomena such
as the quasi-biennial oscillation \citep{lindzen_qbo}. The majority of our paper
is dedicated to characterizing the behavior of the critical layer when
interacting with a continuous train of IGW excited from the bottom of the fluid.
IGW are generally \emph{anti}-diffusive, in that they steepen shear flows
\citep{lindzen_qbo,lecoanet_meanflow}, and act to narrow the critical layer. We
find this steepening is suppressed by the Kelvin-Helmholtz instability and
turbulence within the very narrow critical layer. By careful accounting of the
momentum flux budget about the critical layer, we are able to model the
propagation of the critical layer, reflection of the incident IGW, and other
momentum redistribution behaviors. Our results are numerically converged for
sufficiently large resolutions.

In Section~\ref{s:equations}, we will describe the system of equations we will
use to analyze IGW breaking. In Section~\ref{s:theory}, we review existing
understanding of wave breaking and present new analytical results. In
Section~\ref{s:numerics}, we describe our numerical setup, which in
Section~\ref{s:weak_sim} we validate in the weak forcing limit against linear
theory. In Section~\ref{s:sim}, we present the results of simulation of IGW
breaking and our characterization of the critical layer. Finally, we summarize
and conclude in Section~\ref{s:discussion}.

@@

    \item $\nu$: Nonlinear effects transfer wave energy from the injection
        wavenumber $\bm{k}$ to larger wavenumbers. Our spectral method does not
        have any numerical viscosity, so diffusivity must be introduced into the
        equations to regularize the systems at large wavenumbers. We add
        viscosity and diffusivity to the system in a way that conserves
        horizontal momentum (see Appendix~\ref{se:strat_impl} for details). To
        quantify $\nu$, we define the dimensionless Reynolds number
        \begin{equation}
            \mathrm{Re}  \equiv \frac{\omega}{\nu k_{z}^2}. \label{eq:re_def}
        \end{equation}
        When the forcing is weak, we may set $\nu \approx 0$, or $\mathrm{Re}
        \gg 1$ because there is negligible energy transfer into higher-order
        modes. Note that $\nu = 0$ changes the order of the PDE system, as the
        highest-order derivatives appear only in the dissipitive terms, so a
        nonzero $\nu$ must be chosen even in weakly forced simulations.
\end{enumerate}

@@

\subsection{Physical Sources of Dissipation in WDs}\label{ss:disp}

The most significant linear damping in WD g-modes comes from radiative damping
\citep{fullerI}. In \citep{wu} and \citep{fullerI}, the radiative damping rate
is given in terms of $\omega_i = \gamma \omega_r$, where $\omega_r$ is the
frequency of the g-mode. Typical values for $\gamma$ range from $10^{-4}$ to
$10^{-11}$ depending on $n$ of the g-mode.

We will assume this prescription directly transfers to propagating IGW, which
results in general agreement with \citep{bukart}'s estimate of radiative damping
rates. Then, making coarse identification $\omega_i \sim \nu k^2 \approx \nu
k_z^2$, we find that $\mathrm{Re} \sim \frac{1}{\gamma}$. Even at $\gamma =
10^{-4}$ however, the corresponding $\mathrm{Re}$ is far too weak to suppress
reflection/transmission at the critical layer (Fig.~\ref{fig:agg}).

Another source of dissipation considered in \citep{bukart} is turbulent
conbmtive damping. They find this damping rate to never exceed that of
radiative damping, and so it is also too weak to suppress critical layer
formation in our problem.

Finally, we consider the impact of magnetic winding. In \citep{bukart}, magnetic
winding is used to enforce solid body rotation on the grounds that $t_A \gg
t_{gw}$, where
\begin{equation}
    t_A = \int\limits_0^R \frac{\sqrt{4\pi \rho}}{B_0}\;\mathrm{d}r
        \sim 10^2\;\mathrm{yr}\p{\frac{10^3\;\mathrm{G}}{B}},
\end{equation}
the Alfv\'en wave crossing time (evaluated for a CO WD in \citet{fullerIV}),
measures the magnetic coupling time and $t_{gw}$ measures the gravitational wave
inspiral timescale. Before solid body rotation is attained, another relevant
timescale is the synchronization timescale $t_s$. For a tidal torque $\tau$ and
tidal forcing frequency $\sigma = m\p{\Omega - \Omega_{spin}}$, we note that
angular momentum transfer is $\pd{M_{sync}}{t} \sigma R^2 = \tau$, where
$M_{sync}$ denotes the mass of the WD that has synchronized. Thus, the
synchronization timescale is
\begin{align}
    t_{sync} &\sim \frac{M_{sync}\sigma R^2}{\tau},\\
        &\sim 20\;\mathrm{yr}
            \p{\frac{M_{sync}}{10^{-4}M_{\odot}}}
            \p{\frac{\sigma}{2\pi / (1\;\mathrm{hr})}}
            \p{\frac{R}{R_{\oplus}}}^2
            \p{\frac{10^{-14} GM_{\odot}^2/R_{\oplus}}{\tau}}.
\end{align}
A representative $\tau$ has been taken from \citep{bukart}. Only $M_{sync} \sim
10^{-4}M_{\odot}$ need be heated for the thermodynamically interesting effects
studied in \citep{fullerIV} and \citep{tidal_novae}, it seems that strong shear
flows may exist even in the presence of magnetic coupling.

\subsection{Applicability to Astrophysical Systems}\label{ss:other}

Below, we speculate on the applicability of our results to general astrophysical
problems of interest:
\begin{enumerate}
    \item In astrophysical systems, $k_{\perp} \ll k_r$, while in our study,
        $k_x \simeq k_z$. True turbulence is expected to be isotropic at small
        scales, which may introduce different behavior for different ratios of
        $k_z/k_x$. However, $k_x \ll k_z$ is a numerically difficult regime; we
        defer its exploration to future work.

    \item In~\cite{barker_ogilvie}, inwards-propagating IGW are excited that
        break via geometric focusing. They find no reflected wave despite their
        nonlinear timescales being $10\times$ shorter than their viscous
        timescale, in contrast with the reflective behavior observed in our
        simulations. It is plausible the discrepancy arises due to the differing
        causes of wave breaking (geometric focusing vs.\ density
        stratification), but their simulations are also at significantly higher
        artificial viscosities than ours, as we show below.

        Associating $t_{L} \sim \nu k^2 \approx \nu k_z^2$ with the visocus
        timescale and $t_{NL} \sim \bm{u} \cdot \bm{\nabla} \sim \omega$ for the
        nonlinear timescale, we find their $\lambda = \frac{t_{NL}}{t_L} \sim
        \mathrm{Re}$ our Reynolds number. Our simulations indicate $\mathrm{Re}
        \gtrsim 500$ are required to observe the correct asymptotic behavior in
        terms of horizontal momentum flux reflection/transmission, so it is
        possible their lack of reflection is resolution limited.

    \item Our results give some tentative understanding of where tidal heating
        is expected to occur. While we use a barotropic equation of state and do
        not track heat deposition, we can recall that the strongest small-scale
        variations in fluid flow are localized within the critical layer. Thus,
        we posit that a substantial fraction of the energy carried by the IGW is
        deposited in the very narrow critical layer. Then, as the critical layer
        propagates downwards, the entire star is heated. This tidal heating
        model differs flom that used in \citep{tidal_novae}, and its effects
        will be studied in future work.
\end{enumerate}

>>>>

\title[Physics of Tidal Dissipation]{Physics of Tidal Dissipation in Early-Type
Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave
Breaking in Isothermal Stellar Envelopes}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^2$
Dong Lai,$^{1,3}$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
\\
$^3$ Tsung-Dao Lee Institute, 800 Dongchuan Road, Shanghai, 200240, China
}

@@

\pubyear{2020}

@@

\begin{abstract}
    In binaries composed of early-type stars and white dwarfs, the dominant
    tidal process involve the excitation of internal gravity waves (IGWs) that
    propagate towards the stellar surface and their dissipation via nonlinear
    wave breaking. We perform 2D hydrodynamical simulations of this wave
    breaking process in a stratified isothermal atmosphere. We find that, after
    an initial transient phase, the dissipation of the IGWs naturally generate a
    sharp critical layer, separating the lower stationary region (with no mean
    flow) and the upper ``synchronized'' envelope (with the mean flow speed
    equal to the horizontal wave phase velocity). We find that the indefinite
    steepening of this critical layer is suppressed by the Kelvin-Helmholtz
    instability, and the width of the layer is determined by the Richardson
    criterion. We study the absorption and reflection of incident waves off the
    critical layer and provide analytical formulae describing its evolution. The
    result of this study is important for characterizing the evolution of
    tidally heated white dwarfs and other binary stars.
\end{abstract}

@@

% so that \cite[see][for q]{QQ} gives (see QQ for q) without comma after QQ
\setcitestyle{notesep={ }}

@@

The physical processes responsible for tidal evolution in close binaries often
involve the excitation and dissipation of internal waves, going beyond the
``weak friction'' of equilibrium tides \citep[see][for a
review]{ogilvie2014tidal}. In particular, internal gravity waves (IGWs), arising
from buoyancy of stratified stellar fluid, play an important role in several
types of binary systems. In solar-type stars with radiative cores and convective
envelopes, IGWs are excited by tidal forcing at the radiative-convective
boundary and propagate inward; as the wave amplitude grows due to geometric
focusing, nonlinear effects can lead to efficient damping of the wave
\citep{goodman1998dynamical,barker_ogilvie,essick2015orbital}. In massive
main-sequence stars, with convective cores and radiative envelopes, IGWs are
similarly excited at the convective-radiative interface but travel toward the
stellar surface; nonlinearity develops as the wave amplitude grows, leading to
efficient dissipation \citep{zahn75,zahn77}. As the outgoing wave deposites its
angular momentum to the stellar surface layer, a critical layer may form and the
star synchronize from outside-in \citep{gn89}.

Tidal dissipation can also play an important role in compact double white dwarf
(WD) binary systems (with orbital periods in the range of minutes to hours).
Such binaries may produce a variety of exotic astrophysical systems and
phenomena, ranging from isolated sdB/sdO stars,
%(likely created by the merger of two He WDs),
R CrB stars,
%(likely created by the merger of a CO and He WD),
AM CVn binaries,
%(possibly created by stable mass transfer from a He WD onto a CO WD),
high-mass neutron stars and magnetars (created by the accretion-induced
collapse of merging WDs),
%CO WDs or accreting O-Ne-Mg WDs),
and various optical transients (underluminous supernovae, Ca-rich fast
transients, and type Ia supernovae)
\citep[e.g.][]{livio2018progenitors,gaiaDD2,toloza2019understanding,
underlum,carich,Ia1,Ia2}. The outcomes of WD mergers depend on the WD masses
and composition, but tidal dissipation can strongly affect the pre-merger
conditions of the WDs. Tidal dessipation may also incluence the evoluton of
eccentric WD-massive black hole binaries prior to the eventual tidal disruption
of the WD \citep{vick2017tidal}.

Previous studies have identified nonlinear dissipation of IGWs as the key tidal
process in compact WD binaries \citep{fullerII,fullerIV,tidal_novae,bukart}:
IGWs are tidally excited mainly at the compositon transitions of the WD
envelope; as these waves propagate outwards towards the WD surface, they grow
in amplitude until they break, and transfer both energy and angular momentum
from the binary orbit to the outer envelope of the WD\@.
%Fuller \& Lai (20xx,20xx) found that the dissipation of IGW can generate
%significantly more energy than thermal radiation from the isolated WD surface
%and is thus a major contributor to the WD energy budget.
However, these previous works parameterized the wave breaking process in an ad
hoc manner. The details of dissipation, namely the location and spatial extent
of the wave breaking, affect the observable outcome: Dissipation near the
surface of the WD can be efficiently radiated away and simply brightens the WD,
while dissipation deep in the WD envelope causes an energy buildup that results
in energetic flares \citep{tidal_novae}. An important goal of this paper is to
elucidate the details of the nonlinear IGW breaking process; the result of this
``microphysics'' study will help determine the observational manifestation
tidally heated binary WDs.

In this paper, we perform numerical simulations of IGW breaking in a
plane-parallel stratified atmosphere (a stellar enevelope). We use the
pseudo-spectral code Dedalus \citep{dedalus,dedalus2} and a 2D Cartesian
geometry, and consider IGWs propagating into an isothermal fluid initially at
rest. We find that, after an initial transient phase, a \emph{critical layer}
naturally develops, separating a lower zone that has no horizontal mean flow
and an upper zone with mean flow at the horizontal phase velocity of the IGW\@.
The major part of our paper is dedicated to characterizing the behavior of the
critical layer when interacting with a continuous train of IGW excited from the
bottom of the fluid. IGWs are generally \emph{anti}-diffusive, in that they
steepen shear flows \citep{lindzen_qbo,lecoanet_meanflow} and act to narrow the
critical layer. We find this steepening is suppressed by the Kelvin-Helmholtz
instability and turbulence within the narrow critical layer. By careful
accounting of the momentum flux budget about the critical layer, we are able to
model the reflection and absorption of the incident IGW, and the slow downward
propagation of the critical layer.

While the motivation of our study is to understand tidal dissipation
in WD and mass star binaries, the IGW breaking process studied in this paper
is also quite relevant to the circulation dynamics of planetary atmospheres
\citep[see e.g.][]{lindzen1981turbulence,holton1983influence}.

This paper is organized as follows. In Section~\ref{s:equations} we present the
system of equations used in our simulations. In Section~\ref{s:theory}, we
review the existing understanding of wave breaking and present new analytical
results on IGWs around a critical layer. In Section~\ref{s:numerics} we describe
our numerical setup and in Section~\ref{s:weak_sim} we validate our method in
the weak forcing limit against linear theory. In Section~\ref{s:sim}, we present
the results of simulations of IGW breaking and our characterization of the
critical layer. We summarize and conclude in Section~\ref{s:discussion}.

@@

    \item $\nu$: Nonlinear effects transfer wave energy from the injection
        wavenumber $\bm{k}$ to larger wavenumbers. Our spectral method does not
        have any numerical viscosity, so diffusivity must be introduced into the
        equations to regularize the systems at large wavenumbers. We add
        viscosity and diffusivity to the system in a way that conserves
        horizontal momentum (see Appendix~\ref{se:strat_impl} for details). To
        quantify $\nu$, we define the dimensionless Reynolds number
        \begin{equation}
            \mathrm{Re} \equiv \frac{\omega}{\nu k_{z}^2}
                = \frac{\omega}{\nu}\p{\frac{H}{2\pi}}^2.\label{eq:re_def}
        \end{equation}
        We use $\mathrm{Re} \gg 1$ in our simulations\footenote{This condition
        is always satisfied in stars. For example, in WDs, the dominant linear
        dissipation mechanism of g-modes is radiative damping, with damping rate
        ranging from $10^{-11}$--$10^-4$ of the mode frequency \citep{fullerI}.
        This corresponds to a small effective viscosity or $\mathrm{Re} \gg
        1$.}. Note that $\nu = 0$ changes the order of the PDE system, as the
        highest-order derivatives appear only in the dissipitive terms, so only
        nonzero $\nu$ can be used even in weakly forced simulations.
\end{enumerate}

@@

\subsection{Discussion}

In this paper, we have studied the nonlinear behavior of IGWs with $\abs{k_x /
k_z} \sim 1$ in a plane-parallel atmospheres. On the other hand, tidally
excited IGWs in binary stars have horizontal wave number $k_\perp \sim 1/R$
(where $R$ is the stellar radius) much smaller than the radial wave number
$k_r$. Strictly speaking, our results may not directly apply to IGWs with
$\abs{k_x/k_z}\ll 1$. However the qualitative behavior is likely to be similar,
as the turbulence driving the critical layer dynamics occurs at scales
significantly smaller than either $1/k_x$ or $1/\abs{k_z}$. Simulating IGWs with
$k_x \ll k_z$ is more numerically challenging and we defer its exploration to
future work.

It is interesting to compare our work with that of \citet{barker_ogilvie}, who
studied inward-propagating IGWs in solar-type stars and their nonlinear breaking
due to geometric focusing. In their numerical simulation in a 2D polar geometry,
they find no evidence for reflected waves, contrary to our result. It is
plausible the discrepancy arises due to the differing causes of wave breaking
(geometric focusing vs.\ density rarefaction). Additionally, their nonlinearity
parameter $\lambda \sim 1/\mathrm{Re}$ where $\mathrm{Re}$ is our Reynolds
number (Eq.~\eqref{eq:re_def}). For $\lambda \sim 10^{-1}$ as explored in
\citet{barker_ogilvie}, our results also predict no reflected wave.

Regardless of the limitations inherent in our simulations (e.g.\ plane-parallel
geometry), our results shed light on the physical mechanism of tidal heating in
close binaries. In particular, our simulations indicate that energy dissipation
occurs in a narrow critical layer. The star heats up from outside-in as the
critical layer propagates inwards. This tidal heating profile differs flom that
used by \citet{tidal_novae}. We plan to study this issue in a future work.
